---
title: '(user template) Intro to the Tidyverse: R-Ladies SB May 2019'
author: "Your Name Here"
date: "15 May 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **Don't have the tidyverse yet?**
### Install using the following code:

```{r, eval = FALSE}
install.packages("tidyverse")
```

### Load the tidyverse:

```{r, message = FALSE, warning = FALSE}

```

## **Data wrangling cheat sheet:**

Let's first create some completely hypothetical data about the number of pizzas eaten by Sam, An, Allison, Julie, and Jamie over the past 3 years :)
```{r}

```

We'll first want to transform 'pizza_data' into tidy (long) format:

**gather():** transforms data from wide to long format
```{r}

```

Now let's use our tidy data to practice some further wrangling.

**arrange():** order observations as specified (default = alphabetical or ascending)
```{r}

```

**filter():** select observations within columns
```{r}

```

**group_by():** groups observations such that data operations are performed at the level of the group
```{r}

```

**mutate():** 
```{r}

```

**pull():** pulls out a single variable from a data frame and saves it as a vector
```{r}

```

**rename():** rename a column
```{r}

```

**select():** select columns to retain and specify their order
```{r}

```

**spread():** transform data from long to wide format
```{r}

```

**summarize():** calculate summary statistics
```{r}

```

**tally():** sum values across groups
```{r}

```

## **Now let's practice!**

### Load the tidyverse and any additional required packages:

```{r, message = FALSE, warning = FALSE}

```

### Load the data: 

In celebration of this year's superbloom, we'll be exploring phenometric data of flowering California plants from the [USA -- National Phenology Network](https://www.usanpn.org/home).

```{r, message = FALSE, warning = TRUE}

```

Let's pretend we're trying to plan a getaway to the Joshua Tree National Park and want to time our trip so that we have the greatest chance of seeing fully bloomed plants. 

### Explore: 

We should first familiarize ourselves with the data. 

```{r, eval = FALSE}

```

### Wrangle:

This dataset is *huge*--we'll want to wrangle it so that it only includes the information that we're interested in. We will:

a.   
b. 
c.   
d.   
e.  
f. 
g.

#### a. convert variable names to snake_case using `janitor::clean_names()`

```{r}

```

#### b. filter for California observations

```{r}

```

#### c. select the columns we want

```{r}

```

#### d. rename columns

```{r}

```

#### e. unite columns

```{r}

```

#### f. remove any NA values

```{r}

```

#### g. set the levels for a character vector

```{r}

```

#### Now all together!

```{r}

```

With this simplified and cleaned data set, we're ready to explore a subset of the desert species we're most interested in. We love **Joshua trees** (*Yucca brevifolia*), **creosote bushes** (*Larrea tridentata*), and **Mojave yucca** (*Yucca schidigera*) and want to know when these plants are blooming. Let's first isolate data for these species by:

a. 
b. 
c. 

```{r}

```

### Plot:

a. 
b. 
c. 

```{r, fig.align = 'center', fig.width = 15, fig.height = 10}

```